---
title: "ProjectDraft"
author: "Cam Lunn, Atticus Patrick, & Owen Patrick"
date: "4/18/2022"
output:
  html_document: default
  pdf_document: default
---

|  Name     |       Description                     |    Levels                   |
|:---------:|    :-----------------------------:    |  :------------------------: |
|   Age     |        Age of the patient             |       28 yrs - 77 yrs       |
|   Sex     |        Sex of the patient             |       Male, Female          |
|   exang   |        exercise induced angina        |       (1 = yes; 0 = no)     |
|   caa     |        number of major vessels        |       (0-3)                 |
|   cp      |        Chest Pain type chest pain type|       Value 1: typical angina   [TA] |
|           |                                       |       Value 2: atypical angina  [ATA]|
|           |                                       |       Value 3: non-anginal pain [NAP]|
|           |                                       |       Value 4: asymptomatic     [ASY]|
|   trtbps  |     resting blood pressure (in mm Hg) |       0 - 200 mm Hg                       |
|   chol    |   cholestoral in mg/dl fetched via BMI sensor |   0-603 mg/dl                     |
|   fbs     |        (fasting blood sugar > 120 mg/dl) |        (1 = true; 0 = false)      |   
|   rest_ecg|        resting electrocardiographic results   |  Value 0: normal              |
|           |                                       |          Value 1: having ST-T wave abnormality (T wave inversions and/or ST elevation or  | |           |                                       |          depression of > 0.05 mV)                                                         |
|           |                                       |     Value 2: showing probable or definite left ventricular hypertrophy by Estes'  criteria |
|   thalach |       maximum heart rate achieved     |       60 - 202 bpm                              |
|   target  |        chance of a heart attack       |       0= less chance of heart attack; 1= more chance of heart attack |                                                           

------------------------------------------------------------------------------------------------------------------------------------------------



HEART2: https://www.kaggle.com/datasets/fedesoriano/heart-failure-prediction


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
pacman::p_load(tidyverse, rstatix, class,
               rpart, rpart.plot, dplyr, corrplot, MASS, caret, MVN, 
               factoextra, psych)
source("Partial F Test function.R")

heart2 <- read.csv("heart2.csv")

# ----------------------- Data for Classifying Heart Disease -----------------------

heart <- heart2  %>% 
    mutate(Age = if_else(Age >= 28 & Age <= 37,
                         "28-37",
                         if_else(Age >= 38 & Age <= 47, 
                                 "38-47",
                         if_else(Age >= 48 & Age <= 57, 
                                 "48-57", 
                         if_else(Age >= 58 & Age <= 67, 
                                 "58-67",
                         if_else(Age >= 68 & Age <= 77, 
                                 "68-77", "Not seen"))))), 
           HeartDisease = if_else(HeartDisease == "0",
                       "Unaffected",
                       "Affected") %>% factor())

# ----------------------- Data for Classifying Chest Pain ---------------------- 

heart_CP <- heart2  %>% 
    mutate(Age = if_else(Age >= 28 & Age <= 37,
                         "28-37",
                         if_else(Age >= 38 & Age <= 47, 
                                 "38-47",
                         if_else(Age >= 48 & Age <= 57, 
                                 "48-57", 
                         if_else(Age >= 58 & Age <= 67, 
                                 "58-67",
                         if_else(Age >= 68 & Age <= 77, 
                                 "68-77", "Not seen"))))),
           ChestPainType = factor(ChestPainType,
                      levels = c("TA", "ATA", "NAP", "ASY")))

# Need to edit this?

N <- nrow(heart); p <- ncol(heart %>%dplyr::select(where(is.numeric)));

# number of groups in ChestPainType, k_pain.
k_pain <- n_distinct(heart$ChestPainType)
# number of groups in age, k_age
k_age <- n_distinct(heart$Age)
# number of groups in Heart Disease, k_HD
k_HD <- n_distinct(heart$HeartDisease)


```


```{r Exploratory}
# Combined 5 datasets (Cleveland, Long Beach, Switzerland, Hungarian, & Stalog)
skimr::skim(heart)

R <- cor(heart %>%dplyr::select(where(is.numeric)))

corrplot(R, 
        method="shade", 
        type="upper", 
        addCoef.col = "yellow")

heart %>% 
  pivot_longer(cols = c(RestingBP, Cholesterol, Oldpeak, MaxHR),
               names_to = "attribute",
               values_to = "value") %>% 
  
  ggplot(mapping = aes(x = value,
                       fill = ChestPainType)) + 
  geom_boxplot() + 
  facet_wrap(facets = ~ attribute,
             scales = "free") +
  labs(fill = "ChestPainType") +
  theme(legend.position = "top")

heart %>% 
  pivot_longer(cols = c(RestingBP, Cholesterol, Oldpeak, MaxHR),
               names_to = "attribute",
               values_to = "value") %>% 
  
  ggplot(mapping = aes(x = value,
                       fill = HeartDisease)) + 
  geom_density(alpha = .5) + 
  facet_wrap(facets = ~ attribute,
             scales = "free") +
  labs(fill = "HeartDisease") +
  theme(legend.position = "top")

table(heart$Sex)

```

## Check some group means
```{r Group Means, warning=FALSE}

HD_means <-
  heart %>% 
  group_by(HeartDisease, Age) %>% 
  summarize(across(.cols = c(Cholesterol, Oldpeak, MaxHR),
                   .fns = mean))
view(HD_means)

```

## PCA to Check Significance of Variables 
```{r PCA}

(heart_R_PCA <- prcomp(heart %>% dplyr::select(where(is.numeric)),
                      scale. = T))
summary(heart_R_PCA)


fviz_screeplot(X = heart_R_PCA,
               choice = "eigenvalue",
               geom = "line",
               linecolor = "steelblue",
               ncp = p) + 
  
  labs(title = "Screeplot using the Covariance Matrix",
       x = "Principal Component") + 
  
  geom_hline(yintercept = 1,
             color = "darkred")


```

#### 2c: Correlation Matrix PCA Biplot

```{r 2c_biplot}

fviz_pca(X = heart_R_PCA,
         axes = c(1, 2),              # Which PCs to plot
         geom = c("point"), 
         alpha.ind = .1, 
         repel = T)  +   # text adds name of country.

  coord_equal()


```


## Check Differences Using MANOVA


```{r Manova CHEST PAIN, echo=FALSE}

# heart_man <- manova(cbind(RestingBP, Cholesterol, MaxHR, Oldpeak) ~ Age,
#                   data = heart)
# summary(heart_man)

heart_man <- manova(cbind(RestingBP, Cholesterol, FastingBS, MaxHR, Oldpeak, HeartDisease) ~ ChestPainType,
    data = heart)

summary(heart_man)

```

## Check Assumptions

```{r Assumptions, echo=FALSE}
# Not normal
mvn(data = heart_man$residuals, 
    desc = F, 
    multivariatePlot = "qq",
    univariateTest = "SW",
    mvnTest = "mardia")

box_m(data =  heart[, c(4, 5, 6, 8, 10)],
      group = heart$ChestPainType)

```

## Let's see what's actually useful:

```{r}
Partial_F(Y = heart_CP %>%
              dplyr::select(where(is.numeric)),
          x = heart_CP$ChestPainType)

Partial_F(Y = heart_CP %>% 
              dplyr::select(FastingBS, RestingBP, Cholesterol, Oldpeak, MaxHR, HeartDisease), 
          x = heart_CP$ChestPainType)

heart_man <- manova(cbind(FastingBS, RestingBP, Cholesterol, Oldpeak, MaxHR, HeartDisease) ~ ChestPainType,
    data = heart_CP)

summary(heart_man)
# ------------------------------------------ # 
Partial_F(Y = heart_CP %>% 
              dplyr::select(RestingBP, Cholesterol, Oldpeak, MaxHR, HeartDisease), 
          x = heart_CP$ChestPainType)

heart_man <- manova(cbind(RestingBP, Cholesterol, Oldpeak, MaxHR, HeartDisease) ~ ChestPainType,
    data = heart_CP)

summary(heart_man)

# ------------------------------------------ # 

Partial_F(Y = heart_CP %>% 
              dplyr::select(Cholesterol, Oldpeak, MaxHR, HeartDisease), 
          x = heart_CP$ChestPainType)

heart_man <- manova(cbind(Cholesterol, Oldpeak, MaxHR, HeartDisease) ~ ChestPainType,
    data = heart_CP)

summary(heart_man)

# ------------------------------------------ # 



Partial_F(Y = heart_CP %>% 
              dplyr::select(Oldpeak, MaxHR, HeartDisease), 
          x = heart_CP$ChestPainType)

heart_man <- manova(cbind(Oldpeak, MaxHR, HeartDisease) ~ ChestPainType,
    data = heart_CP)

summary(heart_man)

# ------------------------------------------ # 


Partial_F(Y = heart_CP %>% 
              dplyr::select(MaxHR, HeartDisease), 
          x = heart_CP$ChestPainType)

heart_man <- manova(cbind(MaxHR, HeartDisease) ~ ChestPainType,
    data = heart_CP)

summary(heart_man)


# ------------------------------------------ # 

# --------------CHECKING HEART DISEASE AS OUTCOME ------------------# 

# ------------------------------------------ # 
Partial_F(Y = heart %>% 
              dplyr::select(RestingBP, Cholesterol, Oldpeak, MaxHR), 
          x = heart$HeartDisease)

heart_man <- manova(cbind(RestingBP, Cholesterol, Oldpeak, MaxHR) ~ HeartDisease,
    data = heart)

summary(heart_man)

# ------------------------------------------ # 
Partial_F(Y = heart %>% 
              dplyr::select(Cholesterol, Oldpeak, MaxHR), 
          x = heart$HeartDisease)

heart_man <- manova(cbind(Cholesterol, Oldpeak, MaxHR) ~ HeartDisease,
    data = heart)

summary(heart_man)

# Stratify by ChestPain Type:
heart_man <- manova(cbind(Cholesterol, Oldpeak, MaxHR) ~ ChestPainType + HeartDisease,
    data = heart)

summary(heart_man)

#Age, Sex, RestingBP, Cholesterol, FastingBS, RestingECG, MaxHR, ExerciseAngina, Oldpeak, ST_Slope, HeartDisease
```


```{r Linear Discriminate, echo=FALSE}
# ------------ Plot the discriminant for HEART DISEASE ------------------

heart_HD_lda <- MASS::lda(HeartDisease ~ cbind(Cholesterol, Oldpeak, MaxHR),
                     data = heart)

ld_sep_pct <- round(heart_HD_lda$svd^2/sum(heart_HD_lda$svd^2)*100,
                    digits = 1)

heart_HD <- 
  data.frame(heart, 
             predict(heart_HD_lda)$x)

gg_lda_density <-
  heart_HD %>%
  ggplot(mapping = aes(x = LD1,
                       fill = HeartDisease)) +

  theme(legend.position = "bottom") +

  labs(x = paste0("LD1 (Percent Explained: ", ld_sep_pct[1], "%)"),
       y = paste0("LD2 (Percent Explained: ", ld_sep_pct[2], "%)"))

gg_lda_density +
  geom_density(alpha = .5)

# ------------------ Plot Discrimant for CHEST PAIN -------------------- 
heart_CP_lda <- MASS::lda(ChestPainType ~ cbind(Oldpeak, MaxHR, HeartDisease),
                     data = heart_CP)

ld_sep_pct <- round(heart_CP_lda$svd^2/sum(heart_CP_lda$svd^2)*100,
                    digits = 1)

heart_CPLDA <-
  data.frame(heart_CP, 
             predict(heart_CP_lda)$x)

gg_lda_scatter_CP <-
  heart_CPLDA %>%
  ggplot(mapping = aes(x = LD1,
                       y = LD2,
                       color = ChestPainType)) +

  theme(legend.position = "bottom") +

  labs(x = paste0("LD1 (Percent Explained: ", ld_sep_pct[1], "%)"),
       y = paste0("LD2 (Percent Explained: ", ld_sep_pct[2], "%)"))

gg_lda_scatter_CP +
  geom_point()

```

# QDA Predicting for Heart Disease Outcome

```{r Assumptions HEART DISEASE, echo=FALSE}
# Using best model:
heart_man_HD <- manova(cbind(Cholesterol, Oldpeak, MaxHR) ~ HeartDisease,
    data = heart)


# Not normal
mvn(data = heart_man$residuals, 
    desc = F, 
    multivariatePlot = "qq",
    univariateTest = "SW",
    mvnTest = "mardia")

box_m(data =  heart[, c("Cholesterol", "Oldpeak", "MaxHR")],
      group = heart$HeartDisease)

```

```{r QDA Heart Disease, echo=FALSE}
# Not normal and reject box_m test: 
qda_heart_HD_cv <- MASS::qda(formula = HeartDisease~ cbind(Cholesterol, Oldpeak, MaxHR), 
                      data = heart, 
                      CV = T)

# Confusion Matrix
table(predicted = qda_heart_HD_cv$class, 
      actual = heart$HeartDisease) %>% 
  confusionMatrix()

```

```{r Standardize HD}
# Find the pooled standard deviations:
sd_heart_HD <- 
  summary(heart_man)$SS$Residuals %>%  
  diag() %>% 
  sqrt()/sqrt(N-k_HD)



# Standardize the data using the pooled standard deviations:


# Now we need to divide each variable by the pooled sd:
heart_sc_HD <- 
  scale(heart[, c("Cholesterol", "Oldpeak", "MaxHR")],
        center = T, 
        scale = sd_heart_HD) %>% 
  data.frame()


heart_sc_HD$HeartDisease <- heart$HeartDisease

```



```{r QDA CHEST PAIN, echo=FALSE}
# Not normal
qda_heart_CP_cv <- MASS::qda(formula = ChestPainType ~ cbind(Oldpeak, MaxHR, HeartDisease), 
                      data = heart_CP, 
                      CV = T)

# Confusion Matrix
table(predicted = qda_heart_CP_cv$class, 
      actual = heart_CP$ChestPainType) %>% 
  confusionMatrix()

```

```{r Standardize CHEST PAIN}
# Make Manova model:

heart_man_CP <- manova(cbind(Oldpeak, MaxHR, HeartDisease) ~ ChestPainType,
    data = heart_CP)

# Find the pooled standard deviations:
sd_heart_CP <- 
  summary(heart_man_CP)$SS$Residuals %>%  
  diag() %>% 
  sqrt()/sqrt(N-k_pain)



# Standardize the data using the pooled standard deviations:

typeof(heart_CP$HeartDisease)
# Now we need to divide each variable by the pooled sd:
heart_sc_CP <- 
  scale(heart_CP[, c("Oldpeak", "MaxHR", "HeartDisease")],
        center = T, 
        scale = sd_heart_CP) %>% 
  data.frame()

heart_sc_CP$ChestPainType <- heart_CP$ChestPainType

```

```{r 1eii_kChoice}
## Creating a loop to find the best choice for k
RNGversion("4.0.0")
set.seed(123)
# ---------------------- CHEST PAIN ---------------------- #
sqrt(N/k_pain)
k_choice <-5:27

# data.frame to store the predictions for different choices of k
knn_predictions <- data.frame(Actual = heart_CP$ChestPainType)


# Function knn.cv() performs KNN using cross-validation
# and returns the predicted class based on the nearest neighbors.

# Looping through the different choices of k for knn
for (i in k_choice){

  knn_temp <- class::knn.cv(train = heart_sc_CP %>% dplyr::select(-ChestPainType),
                            cl = heart_sc_CP$ChestPainType,
                            k = i)

  # adding the predicted column to the data set
  knn_predictions <-
    knn_predictions %>%
    add_column(knn_temp)
}

# Changing the column names to better describe the results
colnames(knn_predictions) <- c('Actual', paste0("k", k_choice))

# Calculating the error rate for each choice of k:
knn_predictions %>%
  pivot_longer(cols = starts_with("k"),
               names_to = "k_choice",
               values_to = "prediction") %>%
  group_by(k_choice) %>%
  summarize(incorrect = sum(Actual != prediction),
            positive_rate = mean(Actual == prediction)) %>%
  mutate(k = parse_number(k_choice)) %>%

  ggplot(mapping = aes(x = k,
                       y = positive_rate)) +
  geom_line(color = "darkred",
            size = 1) +

  labs(x = "Choice of k",
       y = "Correct Prediction Percentage") +

  scale_x_continuous(breaks = k_choice) +
  scale_y_continuous(labels = scales::percent)

# k = 15?

# ---------------------- HEART DISEASE ----------------------- #
sqrt(N/k_HD)
k_choice <-11:33

# data.frame to store the predictions for different choices of k
knn_predictions <- data.frame(Actual = heart$HeartDisease)


# Function knn.cv() performs KNN using cross-validation
# and returns the predicted class based on the nearest neighbors.

# Looping through the different choices of k for knn
for (i in k_choice){
  
  knn_temp <- class::knn.cv(train = heart_sc_HD %>% dplyr::select(-HeartDisease), 
                            cl = heart_sc_HD$HeartDisease, 
                            k = i)
  
  # adding the predicted column to the data set
  knn_predictions <- 
    knn_predictions %>% 
    add_column(knn_temp)
}

# Changing the column names to better describe the results
colnames(knn_predictions) <- c('Actual', paste0("k", k_choice))

# Calculating the error rate for each choice of k:
knn_predictions %>% 
  pivot_longer(cols = starts_with("k"),
               names_to = "k_choice",
               values_to = "prediction") %>% 
  group_by(k_choice) %>% 
  summarize(incorrect = sum(Actual != prediction),
            positive_rate = mean(Actual == prediction)) %>% 
  mutate(k = parse_number(k_choice)) %>% 
  
  ggplot(mapping = aes(x = k,
                       y = positive_rate)) +
  geom_line(color = "darkred",
            size = 1) + 
  
  labs(x = "Choice of k",
       y = "Correct Prediction Percentage") +
  
  scale_x_continuous(breaks = k_choice) +
  scale_y_continuous(labels = scales::percent)

# k = 19?


```


```{r 1eii_kNNCM}
# ---------------------- CHEST PAIN ---------------------- # 
# # Best choice of kNN model
heart_knn <- knn.cv(train = heart_sc_CP%>% dplyr::select(-ChestPainType),
                          cl = heart_sc_CP$ChestPainType,
                          k = 15)

# Confusion matrix
data.frame(actual = heart_CP$ChestPainType,
           predicted = heart_knn) %>%
  table() %>%
  confusionMatrix()


# ---------------------- HEART DISEASE ---------------------- #

# Best choice of kNN model
heart_knn <- knn.cv(train = heart_sc_HD %>% dplyr::select(-HeartDisease), 
                          cl = heart_sc_HD$HeartDisease, 
                          k = 19)

# Confusion matrix 
data.frame(actual = heart$HeartDisease, 
           predicted = heart_knn) %>%
  table() %>%
  confusionMatrix()



```


```{r Tree CHEST PAIN, message=FALSE}
# Include the two lines below at the top of the R code to ensure your answer matches the solutions
RNGversion("4.0.0")
set.seed(123)
typeof(heart_CP$HeartDisease)

# Create the full classification tree
heart_tree_CP <- rpart(ChestPainType ~ ., 
                   data = heart_CP, 
                   minsplit = 2, 
                   minbucket = 1, 
                   cp = -1,
                   method = "class")
rpart.plot(heart_tree_CP, type=5)


# Looking at the cp table to find the optimal pruning value:
# simplest tree where xerror < min(xerror) + min(xstd)
printcp(heart_tree_CP)
plotcp(heart_tree_CP)

```

```{r CHEST PAIN Prune Tree}
# Prune the tree

p_heart_tree_CP<- prune(heart_tree_CP, cp= 0.046)


# Plot the pruned tree

rpart.plot(p_heart_tree_CP, 
           type=5, 
           extra = 101)


# Display the confusion matrix
pheart_tree_pred <- predict(object = p_heart_tree_CP,
                         newdata = heart_CP, 
                         type = 'class')

data.frame(actual = heart_CP$ChestPainType, 
           predicted = pheart_tree_pred) %>%
  table() %>%
  confusionMatrix()

```


``` {r HEART DISEASE Tree}

# Include the two lines below at the top of the R code to ensure your answer matches the solutions
RNGversion("4.0.0")
set.seed(123)


# Create the full classification tree
heart_tree2 <- rpart(HeartDisease ~ .,
                   data = heart,
                   minsplit = 2,
                   minbucket = 1,
                   cp = -1,
                   method = "class")
rpart.plot(heart_tree2, type=5)


# Looking at the cp table to find the optimal pruning value:
# simplest tree where xerror < min(xerror) + min(xstd)
printcp(heart_tree2)
plotcp(heart_tree2)



# Prune the tree

p_heart_tree2<- prune(heart_tree2, cp= 0.0182927)


# Plot the pruned tree

rpart.plot(p_heart_tree2,
           type=5,
           extra = 101)


# Display the confusion matrix
pheart_tree_pred2 <- predict(object = p_heart_tree2,
                         newdata = heart,
                         type = 'class')

data.frame(actual = heart$HeartDisease,
           predicted = pheart_tree_pred2) %>%
  table() %>%
  confusionMatrix()


```



