---
title: "ProjectDraft"
author: "Cam Lunn, Atticus Patrick, and Owen Patrick"
date: "4/18/2022"
output: html_document
---
1. Age : Age of the patient
2. Sex : Sex of the patient
3. exang: exercise induced angina (1 = yes; 0 = no)
4. caa: number of major vessels (0-3)
5. cp : Chest Pain type chest pain type
  Value 1: typical angina   [TA]
  Value 2: atypical angina  [ATA]
  Value 3: non-anginal pain [NAP]
  Value 4: asymptomatic     [ASY]
6. trtbps : resting blood pressure (in mm Hg)
7. chol : cholesterol in mg/dl fetched via BMI sensor
8. fbs : (fasting blood sugar > 120 mg/dl) (1 = true; 0 = false)
9. rest_ecg : resting electrocardiographic results
  Value 0: normal
  Value 1: having ST-T wave abnormality (T wave inversions and/or ST elevation or depression of > 0.05 mV)
  Value 2: showing probable or definite left ventricular hypertrophy by Estes' criteria
10. thalach : maximum heart rate achieved
11. target : 0= less chance of heart attack; 1= more chance of heart attack


HEART2: https://www.kaggle.com/datasets/fedesoriano/heart-failure-prediction


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
pacman::p_load(tidyverse, rstatix, class,
               rpart, rpart.plot, dplyr, corrplot, MASS, caret, MVN, 
               factoextra)
source("Partial F Test function.R")

# heart <- read.csv("heart.csv")
heart2 <- read.csv("heart2.csv")
table(heart2$HeartDisease)
#as.character(heart2$Age)
heart <- heart2  %>% 
    mutate(Age = if_else(Age >= 28 & Age <= 37,
                         "28-37",
                         if_else(Age >= 38 & Age <= 47, 
                                 "38-47",
                         if_else(Age >= 48 & Age <= 57, 
                                 "48-57", 
                         if_else(Age >= 58 & Age <= 67, 
                                 "58-67",
                         if_else(Age >= 68 & Age <= 77, 
                                 "68-77", "Not seen"))))), 
           HeartDisease = if_else(HeartDisease == "0",
                       "Unaffected",
                       "Affected") %>% factor())

# FastingBS = factor(FastingBS, 
#                        labels = c("Low Levels",
#                                   "High Levels"))

table(heart$HeartDisease)
table(heart$FastingBS)

N <- nrow(heart); p <- ncol(heart %>%dplyr::select(where(is.numeric)));

# number of groups in ChestPainType, k_pain.
k_pain <- n_distinct(heart$ChestPainType)
# number of groups in age, k_age
k_age <- n_distinct(heart$Age)
# number of groups in Heart Disease, k_HD
k_HD <- n_distinct(heart$HeartDisease)

table(heart$Sex)

```


```{r Exploratory}
# Just Cleveland
#skimr::skim(heart)

# Combined 5 datasets (Cleveland, Long Beach, Switzerland, Hungarian, & Stalog)
skimr::skim(heart)

R <- cor(heart %>%dplyr::select(where(is.numeric)))

corrplot(R, 
        method="shade", 
        type="upper", 
        addCoef.col = "yellow")

heart %>% 
  pivot_longer(cols = where(is.numeric),
               names_to = "attribute",
               values_to = "value") %>% 
  
  ggplot(mapping = aes(x = value,
                       fill = ChestPainType)) + 
  geom_boxplot() + 
  facet_wrap(facets = ~ attribute,
             scales = "free") +
  labs(fill = "ChestPainType") +
  theme(legend.position = "top")

heart %>% 
  pivot_longer(cols = c(FastingBS, RestingBP, Cholesterol, Oldpeak, MaxHR),
               names_to = "attribute",
               values_to = "value") %>% 
  
  ggplot(mapping = aes(x = value,
                       fill = HeartDisease)) + 
  geom_boxplot() + 
  facet_wrap(facets = ~ attribute,
             scales = "free") +
  labs(fill = "HeartDisease") +
  theme(legend.position = "top")

table(heart$Sex)

```

## PCA to Check Significance of Variables 
```{r PCA}

(heart_R_PCA <- prcomp(heart %>% dplyr::select(where(is.numeric)),
                      scale. = T))
summary(heart_R_PCA)


fviz_screeplot(X = heart_R_PCA,
               choice = "eigenvalue",
               geom = "line",
               linecolor = "steelblue",
               ncp = p) + 
  
  labs(title = "Screeplot using the Covariance Matrix",
       x = "Principal Component") + 
  
  geom_hline(yintercept = 1,
             color = "darkred")


```

#### 2c: Correlation Matrix PCA Biplot

```{r 2c_biplot}

fviz_pca(X = heart_R_PCA,
         axes = c(1, 2),              # Which PCs to plot
         geom = c("point"), 
         alpha.ind = .1, 
         repel = T)  +   # text adds name of country.

  coord_equal()


```


## Check Differences Using MANOVA


```{r Manova, echo=FALSE}

# heart_man <- manova(cbind(RestingBP, Cholesterol, MaxHR, Oldpeak) ~ Age,
#                   data = heart)
# summary(heart_man)

heart_man <- manova(cbind(RestingBP, Cholesterol, FastingBS, MaxHR, Oldpeak, HeartDisease) ~ Age + ChestPainType,
    data = heart)

summary(heart_man)

```

## Check Assumptions

```{r Assumptions, echo=FALSE}
# Not normal
mvn(data = heart_man$residuals, 
    desc = F, 
    multivariatePlot = "qq",
    univariateTest = "SW",
    mvnTest = "mardia")

box_m(data =  heart[, c(4, 5, 6, 8, 10)],
      group = heart$ChestPainType)

```

## Let's see what's actually useful:

```{r}
# Partial_F(Y = heart %>% 
#               dplyr::select(where(is.numeric)), 
#           x = heart$ChestPainType)

Partial_F(Y = heart %>% 
              dplyr::select(FastingBS, RestingBP, Cholesterol, Oldpeak, MaxHR, HeartDisease), 
          x = heart$ChestPainType)

heart_man <- manova(cbind(FastingBS, RestingBP, Cholesterol, Oldpeak, MaxHR, HeartDisease) ~ ChestPainType,
    data = heart)

summary(heart_man)

# ------------------------------------------ # 
Partial_F(Y = heart %>% 
              dplyr::select(RestingBP, Cholesterol, Oldpeak, MaxHR, HeartDisease), 
          x = heart$ChestPainType)

heart_man <- manova(cbind(RestingBP, Cholesterol, Oldpeak, MaxHR, HeartDisease) ~ ChestPainType,
    data = heart)

summary(heart_man)

# ------------------------------------------ # 

Partial_F(Y = heart %>% 
              dplyr::select(Cholesterol, Oldpeak, MaxHR, HeartDisease), 
          x = heart$ChestPainType)

heart_man <- manova(cbind(Cholesterol, Oldpeak, MaxHR, HeartDisease) ~ ChestPainType,
    data = heart)

summary(heart_man)

# ------------------------------------------ # 



Partial_F(Y = heart %>% 
              dplyr::select(Oldpeak, MaxHR, HeartDisease), 
          x = heart$ChestPainType)

heart_man <- manova(cbind(Oldpeak, MaxHR, HeartDisease) ~ ChestPainType,
    data = heart)

summary(heart_man)

# ------------------------------------------ # 


Partial_F(Y = heart %>% 
              dplyr::select(MaxHR, HeartDisease), 
          x = heart$ChestPainType)

heart_man <- manova(cbind(MaxHR, HeartDisease) ~ ChestPainType,
    data = heart)

summary(heart_man)


# ------------------------------------------ # 

# --------------CHECKING HEART DISEASE AS OUTCOME ------------------# 

# ------------------------------------------ # 
Partial_F(Y = heart %>% 
              dplyr::select(RestingBP, Cholesterol, Oldpeak, MaxHR), 
          x = heart$HeartDisease)

heart_man <- manova(cbind(RestingBP, Cholesterol, Oldpeak, MaxHR) ~ HeartDisease,
    data = heart)

summary(heart_man)

# ------------------------------------------ # 
Partial_F(Y = heart %>% 
              dplyr::select(Cholesterol, Oldpeak, MaxHR), 
          x = heart$HeartDisease)

heart_man <- manova(cbind(Cholesterol, Oldpeak, MaxHR) ~ HeartDisease,
    data = heart)

summary(heart_man)

# Stratify by ChestPain Type:
heart_man <- manova(cbind(Cholesterol, Oldpeak, MaxHR) ~ ChestPainType + HeartDisease,
    data = heart)

summary(heart_man)

#Age, Sex, RestingBP, Cholesterol, FastingBS, RestingECG, MaxHR, ExerciseAngina, Oldpeak, ST_Slope, HeartDisease
```

```{r Group Means}

HD_means <-
  heart %>% 
  group_by(HeartDisease, ChestPainType) %>% 
  summarize(across(.cols = c(Cholesterol, Oldpeak, MaxHR),
                   .fns = mean))
view(HD_means)

```

```{r Linear Discriminate, echo=FALSE}

# heart_lda <- MASS::lda(ChestPainType ~ cbind(Oldpeak, MaxHR, HeartDisease),
#                      data = heart)

heart_lda <- MASS::lda(HeartDisease ~ cbind(Cholesterol, Oldpeak, MaxHR),
                     data = heart)

# heart_lda <- MASS::lda(HeartDisease ~ ., 
#                      data = heart)
# Plot the discriminant

ld_sep_pct <- round(heart_lda$svd^2/sum(heart_lda$svd^2)*100,
                    digits = 1)

heart2 <- 
  data.frame(heart, 
             predict(heart_lda)$x)

gg_lda_density <-
  heart2 %>%
  ggplot(mapping = aes(x = LD1,
                       fill = HeartDisease)) +

  theme(legend.position = "bottom") +

  labs(x = paste0("LD1 (Percent Explained: ", ld_sep_pct[1], "%)"),
       y = paste0("LD2 (Percent Explained: ", ld_sep_pct[2], "%)"))

gg_lda_density +
  geom_density(alpha = .5)

# gg_lda_scatter <- 
#   heart2 %>% 
#   ggplot(mapping = aes(x = LD1, 
#                        y = LD2,
#                        color = ChestPainType)) + 
#   
#   theme(legend.position = "bottom") + 
#   
#   labs(x = paste0("LD1 (Percent Explained: ", ld_sep_pct[1], "%)"),
#        y = paste0("LD2 (Percent Explained: ", ld_sep_pct[2], "%)"))
# 
# gg_lda_scatter + 
#   geom_point()

```

# QDA Predicting for Heart Disease Outcome

```{r Assumptions, echo=FALSE}
# Using best model:
heart_man <- manova(cbind(Cholesterol, Oldpeak, MaxHR) ~ HeartDisease,
    data = heart)

# Not normal
mvn(data = heart_man$residuals, 
    desc = F, 
    multivariatePlot = "qq",
    univariateTest = "SW",
    mvnTest = "mardia")

box_m(data =  heart[, c("Cholesterol", "Oldpeak", "MaxHR")],
      group = heart$HeartDisease)

```

```{r QDA Heart Disease, echo=FALSE}
# Not normal and reject box_m test: 
qda_heart_cv <- MASS::qda(formula = HeartDisease~ cbind(Cholesterol, Oldpeak, MaxHR), 
                      data = heart, 
                      CV = T)

# Confusion Matrix
table(predicted = qda_heart_cv$class, 
      actual = heart$HeartDisease) %>% 
  confusionMatrix()

```

```{r Standardize HD}
# Find the pooled standard deviations:
sd_heart <- 
  summary(heart_man)$SS$Residuals %>%  
  diag() %>% 
  sqrt()/sqrt(N-k_HD)



# Standardize the data using the pooled standard deviations:


# Now we need to divide each variable by the pooled sd:
heart_sc <- 
  scale(heart[, c("Cholesterol", "Oldpeak", "MaxHR")],
        center = T, 
        scale = sd_heart) %>% 
  data.frame()


heart_sc$HeartDisease <- heart$HeartDisease

```



```{r QDA, echo=FALSE}
# Not normal
qda_heart_cv <- MASS::qda(formula = ChestPainType ~ cbind(Oldpeak, MaxHR, HeartDisease), 
                      data = heart, 
                      CV = T)

# Confusion Matrix
table(predicted = qda_heart_cv$class, 
      actual = heart$ChestPainType) %>% 
  confusionMatrix()

```

```{r Standardize}
# Find the pooled standard deviations:
sd_heart <- 
  summary(heart_man)$SS$Residuals %>%  
  diag() %>% 
  sqrt()/sqrt(N-k_pain)



# Standardize the data using the pooled standard deviations:


# Now we need to divide each variable by the pooled sd:
heart_sc <- 
  scale(heart %>% dplyr::select(where(is.numeric)),
        center = T, 
        scale = sd_heart) %>% 
  data.frame()

heart_sc$ChestPainType <- heart$ChestPainType

```

```{r 1eii_kChoice}
## Creating a loop to find the best choice for k
RNGversion("4.0.0")
set.seed(123)
# ---------------------- CHEST PAIN ---------------------- #
# sqrt(N/k_pain)
# k_choice <-5:27
# 
# # data.frame to store the predictions for different choices of k
# knn_predictions <- data.frame(Actual = heart$ChestPainType)
# 
# 
# # Function knn.cv() performs KNN using cross-validation
# # and returns the predicted class based on the nearest neighbors.
# 
# # Looping through the different choices of k for knn
# for (i in k_choice){
#   
#   knn_temp <- class::knn.cv(train = heart_sc %>% dplyr::select(-ChestPainType), 
#                             cl = heart_sc$ChestPainType, 
#                             k = i)
#   
#   # adding the predicted column to the data set
#   knn_predictions <- 
#     knn_predictions %>% 
#     add_column(knn_temp)
# }
# 
# # Changing the column names to better describe the results
# colnames(knn_predictions) <- c('Actual', paste0("k", k_choice))
# 
# # Calculating the error rate for each choice of k:
# knn_predictions %>% 
#   pivot_longer(cols = starts_with("k"),
#                names_to = "k_choice",
#                values_to = "prediction") %>% 
#   group_by(k_choice) %>% 
#   summarize(incorrect = sum(Actual != prediction),
#             positive_rate = mean(Actual == prediction)) %>% 
#   mutate(k = parse_number(k_choice)) %>% 
#   
#   ggplot(mapping = aes(x = k,
#                        y = positive_rate)) +
#   geom_line(color = "darkred",
#             size = 1) + 
#   
#   labs(x = "Choice of k",
#        y = "Correct Prediction Percentage") +
#   
#   scale_x_continuous(breaks = k_choice) +
#   scale_y_continuous(labels = scales::percent)

# k = 21?

# ---------------------- HEART DISEASE ----------------------- #
sqrt(N/k_HD)
k_choice <-11:33

# data.frame to store the predictions for different choices of k
knn_predictions <- data.frame(Actual = heart$HeartDisease)


# Function knn.cv() performs KNN using cross-validation
# and returns the predicted class based on the nearest neighbors.

# Looping through the different choices of k for knn
for (i in k_choice){
  
  knn_temp <- class::knn.cv(train = heart_sc %>% dplyr::select(-HeartDisease), 
                            cl = heart_sc$HeartDisease, 
                            k = i)
  
  # adding the predicted column to the data set
  knn_predictions <- 
    knn_predictions %>% 
    add_column(knn_temp)
}

# Changing the column names to better describe the results
colnames(knn_predictions) <- c('Actual', paste0("k", k_choice))

# Calculating the error rate for each choice of k:
knn_predictions %>% 
  pivot_longer(cols = starts_with("k"),
               names_to = "k_choice",
               values_to = "prediction") %>% 
  group_by(k_choice) %>% 
  summarize(incorrect = sum(Actual != prediction),
            positive_rate = mean(Actual == prediction)) %>% 
  mutate(k = parse_number(k_choice)) %>% 
  
  ggplot(mapping = aes(x = k,
                       y = positive_rate)) +
  geom_line(color = "darkred",
            size = 1) + 
  
  labs(x = "Choice of k",
       y = "Correct Prediction Percentage") +
  
  scale_x_continuous(breaks = k_choice) +
  scale_y_continuous(labels = scales::percent)

# k = 19?


```


```{r 1eii_kNNCM}
# ---------------------- CHEST PAIN ---------------------- # 
# # Best choice of kNN model
# heart_knn <- knn.cv(train = heart_sc %>% dplyr::select(-ChestPainType), 
#                           cl = heart_sc$ChestPainType, 
#                           k = 21)
# 
# # Confusion matrix 
# data.frame(actual = heart$ChestPainType, 
#            predicted = heart_knn) %>%
#   table() %>%
#   confusionMatrix()


# ---------------------- HEART DISEASE ---------------------- #

# Best choice of kNN model
heart_knn <- knn.cv(train = heart_sc %>% dplyr::select(-HeartDisease), 
                          cl = heart_sc$HeartDisease, 
                          k = 19)

# Confusion matrix 
data.frame(actual = heart$HeartDisease, 
           predicted = heart_knn) %>%
  table() %>%
  confusionMatrix()



```


```{r 1eiii_fullTree, message=false}
# Include the two lines below at the top of the R code to ensure your answer matches the solutions
RNGversion("4.0.0")
set.seed(123)


# Create the full classification tree
heart_tree <- rpart(ChestPainType ~ ., 
                   data = heart, 
                   minsplit = 2, 
                   minbucket = 1, 
                   cp = -1,
                   method = "class")
rpart.plot(heart_tree, type=5)


# Looking at the cp table to find the optimal pruning value:
# simplest tree where xerror < min(xerror) + min(xstd)
printcp(heart_tree)
plotcp(heart_tree)
# cp = 0.01236264 is best place to split (0.58516 < 0.56044 + 0.035217)

```

```{r 1eiii_pruneTree}
# Prune the tree

p_heart_tree<- prune(heart_tree, cp= 0.04502370)


# Plot the pruned tree

rpart.plot(p_heart_tree, 
           type=5, 
           extra = 101)


# Display the confusion matrix
pheart_tree_pred <- predict(object = p_heart_tree,
                         newdata = heart, 
                         type = 'class')

data.frame(actual = heart$ChestPainType, 
           predicted = pheart_tree_pred) %>%
  table() %>%
  confusionMatrix()

```







